First Order Alternation

### 2017-08-16 quantifiers/draft.tex
ok

\begin{proposition}\label{prop:substitution}
  Given a formula $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$, where $Q_1 \ldots
  Q_n \in \set{\exists,\forall}$ and $\phi$ is a quantifier-free
  formula with free variables $\fv{}{\phi} = \set{x_1,\ldots,x_n,q}$,
  for any formula $\psi$, the following equivalence holds:
  \[Q_1 x_1 \ldots Q_n x_n ~.~ \phi[\psi/q] \iff 
  Q_1 x_1 \ldots Q_n x_n \exists q ~.~ \phi \wedge (q \rightarrow \psi)\]
\end{proposition}
\proof{

\qed}

>>>>

\begin{proposition}\label{prop:substitution}
  Given a formula $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$, where $Q_1 \ldots
  Q_n \in \set{\exists,\forall}$ and $\phi$ is a quantifier-free
  formula with free variables $\fv{}{\phi} = \set{x_1,\ldots,x_n,q}$
  in which the boolean variable $q$ occurs positively, for any formula
  $\psi$, the following equivalence holds:
  \[Q_1 x_1 \ldots Q_n x_n ~.~ \phi[\psi/q] \iff 
  Q_1 x_1 \ldots Q_n x_n \exists q ~.~ \phi \wedge (q \rightarrow \psi)\]
\end{proposition}
\proof{ Let $\nu$ be an arbitrary valuation of the variables in
  $\fv{}{\phi} \cup \fv{}{\psi}$. It is sufficient to prove that $\nu
  \models \phi[\psi/q]$ if and only if $\nu \models \exists q ~.~ \phi
  \wedge (q \rightarrow \psi)$.

  ``$\Leftarrow$'' Suppose that $\nu \models \phi[\psi/q]$ and
  distinguish the following cases: \begin{compactenum}
  \item if $\nu \models \psi$, since $q$ occurs positively in $\phi$, we
    obtain that $\nu \models \phi[\top/q]$, thus $\nu \models
    \phi[\top/q] \wedge (\top \rightarrow \psi)$ and $\nu \models
    \exists q ~.~ \phi \wedge (q \rightarrow \psi)$ follows.
    %
  \item if $\nu \not\models \psi$, since $q$ occurs positively in $\phi$, we
    obtain that $\nu \models \phi[\bot/q]$, thus $\nu \models
    \phi[\bot/q] \wedge (\bot \rightarrow \psi)$ and $\nu \models
    \exists q ~.~ \phi \wedge (q \rightarrow \psi)$ follows.
  \end{compactenum}

  ``$\Rightarrow$'' Suppose that $\nu \models \exists q ~.~ \phi
  \wedge (q \rightarrow \psi)$, or equivalently, $\nu \models
  (\phi[\top/q] \wedge \psi) \vee \phi[\bot/q]$. If $\nu \models
  \phi[\top/q] \wedge \psi$ then $\nu \models
  \phi[\psi/q]$. Otherwise, if $\nu \models \phi[\bot/q]$, we
  distinguish two cases: \begin{compactitem} 
  \item if $\nu \not\models \psi$ then clearly $\nu \models
    \phi[\psi/q]$.
    %
  \item else, if $\nu \models \psi$, since $q$ occurs positively in
    $\phi$ and $\nu \models \phi[\bot/q]$ then also $\nu \models
    \phi[\top/q]$, because $\phi$ is monotone in $q$ in the entailment
    partial order. But then $\nu \models \phi[\psi/q]$.
  \end{compactitem}
  \qed}
